---
title: "QC analysis of archival samples"
author: "Ludvig Larsson"
date: "11/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)
library(ggplot2)
library(magick)
library(gprofiler2)
library(dplyr)

```


## Prep data
***

Assemble output files from spaceranger and load into a Seurat object.

Sample information:

V10F24-111_A1 = archival sample stored for 37 years
V10F24-111_B1 = archival sample stored for 18 years
V10F24-111_C1 = archival sample stored for 10 years
V10F24-111_D1 = archival sample stored for 5 years

```{r read_input}

samples <- list.files(path = "../data/archivalcancer/", pattern = "filtered_feature", recursive = TRUE, full.names = TRUE)
imgs <- list.files(path = "../data/archivalcancer/", pattern = "hires_image", recursive = TRUE, full.names = TRUE)
spotfiles <- list.files(path = "../data/archivalcancer/", pattern = "tissue_positions", recursive = TRUE, full.names = TRUE)
json <- list.files(path = "../data/archivalcancer/", pattern = "scalefactors", recursive = TRUE, full.names = TRUE)

infoTable <- data.frame(samples, imgs, spotfiles, json, 
                        age = c("37y", "18y", "10y", "5y"),
                        sample_id = paste0("V10F24-111_", c("A", "B", "C", "D"), "1"),
                        stringsAsFactors = FALSE)

se <- InputFromTable(infoTable)

```

## Quality control
***

Summarize per sample quality metrics.

```{r qc}

QC <- function(object) {
  smpls <- GetStaffli(object)@meta.data$sample
  unique.smpls <- unique(smpls)
  df <- do.call(rbind, lapply(unique.smpls, function(s) {
    m <- object@assays$RNA@counts[, smpls == s]
    m <- data.frame(nGene = colSums(m > 0), nUMI = colSums(m), sample = s, stringsAsFactors = FALSE)
  }))
  df %>% group_by(sample) %>%
    summarize(mean_nGene = mean(nGene), median_nGene = median(nGene), max_nGene = max(nGene), min_nGene = min(nGene), 
              mean_nUMI = mean(nUMI), median_nUMI = median(nUMI), max_nUMI = max(nUMI), min_nUMI = min(nUMI))
}

qc.df <- QC(se)
qc.df

```

Visualize quality metric distributions per sample: nFeature_RNA, nCount_RNA percent mitochondrial genes and percent ribosomal protein coding genes

```{r fig.width=12, fig.height=5}

se$age <- factor(se$age, levels = c("37y", "18y", "10y", "5y"))
VlnPlot(se, features = c("nFeature_RNA", "nCount_RNA"), group.by = "age", pt.size = 0.5)

mt.genes <- grep(pattern = "^MT-", x = rownames(se), value = T)
ribo.genes <- grep(pattern = "^RPL|^RPS", x = rownames(se), value = T)
se$percent.mito <- Matrix::colSums(se@assays$RNA@counts[mt.genes, ])/se$nFeature_RNA
se$percent.ribo <- Matrix::colSums(se@assays$RNA@counts[ribo.genes, ])/se$nFeature_RNA

VlnPlot(se, features = c("percent.mito", "percent.ribo"), group.by = "age", pt.size = 0.5) & scale_y_continuous(labels = scales::percent)

```

## Date
***

```{r date}
date()
```

## Session info
***

```{r session}
sessionInfo()

```
