---
title: "mouse brain multiple sections"
author: "Ludvig Larsson"
date: "12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

NOTE: Make sure that you have sctransform v0.2.1 installed to reproduce the plots. Later versions might change the normalization procedure and thereby the analyses.

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)
library(harmony)

```

# Load data
***

Here we have included 7 fresh frozen (FF) and 7 FFPE mouse brain section datasets for the analysis.

5 FF datasets needs to be downloaded from the 10x genomcis website: https://www.10xgenomics.com/resources/datasets

These are named:
Mouse Brain Section (Coronal)
Mouse Brain Serial Section 1 (Sagittal-Anterior)
Mouse Brain Serial Section 1 (Sagittal-Posterior)
Mouse Brain Serial Section 2 (Sagittal-Anterior)
Mouse Brain Serial Section 2 (Sagittal-Posterior)

To be able to run this notebook you will have to download these datasets to the data/mousebrain directory of 
this repo. You will also need to make sure that the datasets are ordered according to the "sample_id" column 
of infoTable. Here's how the different datasets were named:

FF1 = Mouse Brain Section (Coronal)
FF2 = Mouse Brain Serial Section 1 (Sagittal-Anterior)
FF3 = Mouse Brain Serial Section 1 (Sagittal-Posterior)
FF4 = Mouse Brain Serial Section 2 (Sagittal-Anterior)
FF5 = Mouse Brain Serial Section 2 (Sagittal-Posterior)

```{r read_input}

samples <- list.files(path = "../data/mousebrain/", pattern = "filtered_feature_bc_matrix.h5", recursive = T, full.names = T)
imgs <- list.files(path = "../data/mousebrain/", pattern = "tissue_hires_image.png", recursive = T, full.names = T)
spotfiles <- list.files(path = "../data/mousebrain/", pattern = "tissue_positions_list.csv", recursive = T, full.names = T)
json <- list.files(path = "../data/mousebrain/", pattern = "scalefactors_json.json", recursive = T, full.names = T)

infoTable <- data.frame(samples, imgs, spotfiles, json, 
                        condition = c("normal", "normal", "normal", "normal", "normal", "normal", "normal", "normal", "normal", "normal", "normal",  "RU", "RU", "normal"),
                        date = c("NA", "NA", "NA", "NA", "NA", "NA", "NA", "200925", "200925", "200925", "200925", "201016", "201016", "200513"), 
                        sample_id = c("FF1", "FF2", "FF3", "FF4", "FF5", "FF6", "FF7", "FFPE1", "FFPE2", "FFPE3", "FFPE4", "FFPE5", "FFPE6", "FFPE7"), 
                        stringsAsFactors = F)

se <- InputFromTable(infoTable)

```

Get gene attributes

```{r, fig.width=12, fig.height=6}

# Load annotation table with gene biotype information
ensids <- read.table("../data/annotation/mm10_genes.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(ensids) <- ensids$gene_name

gene_attr <- do.call(cbind, lapply(unique(se$sample_id), function(i) {
  count = Matrix::rowSums(se@assays$RNA@counts[, se$sample_id %in% i])
}))
colnames(gene_attr) <- unique(se$sample_id)
gene_attr <- data.frame(gene = rownames(se), biotype = ensids[rownames(se), ]$gene_biotype, gene_attr)
gene_attr_melt <- reshape2::melt(gene_attr, measure.vars = colnames(gene_attr)[-c(1, 2)])

# mt-genes
mt.genes <- grep(pattern = "^mt-", x = gene_attr_melt$gene, value = T)
gene_attr_melt$biotype <- ifelse(gene_attr_melt$gene %in% mt.genes, "protein_coding_mitochondrial", gene_attr_melt$biotype)

# ribo-genes
ribo.genes <- grep(pattern = "Rpl|Rps", x = gene_attr_melt$gene, value = T)
gene_attr_melt$biotype <- ifelse(gene_attr_melt$gene %in% ribo.genes, "protein_coding_ribosomal", gene_attr_melt$biotype)

gene_attr.summarized <- gene_attr_melt %>% 
  dplyr::group_by(biotype, variable) %>%
  dplyr::summarize(count = sum(value)) %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(count = count/sum(count))

gene_attr.summarized$condition <- gsub(pattern = "[1-9]", replacement = "", x = gene_attr.summarized$variable)
gene_attr.summarized$variable <- factor(gene_attr.summarized$variable, levels = c(paste0("FF", 1:7), paste0("FFPE", 1:7)))
p <- ggplot(gene_attr.summarized, aes(x = variable, y = count, fill = biotype)) +
  geom_bar(width = 0.9, stat = "identity", color = "black") +
  theme_classic() +
  labs(y = "RNA biotype content") +
  facet_grid(~condition, scales = "free_x", space = "free_x") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#332288", "#6699CC", "#88CCEE", "#44AA99", "#117733", "#999933", "#DDCC77", "#661100", "#CC6677", "#AA4466", "#882255", "#AA4499", "orange"))

p

```


# date
***

```{r date}
date()
```


# Session info
***

```{r session}
sessionInfo()
```

