---
title: "Analysis of HGSC tumors"
author: "Ludvig Larsson"
date: "5/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)
library(ggplot2)
library(magick)
library(gprofiler2)
library(dplyr)

```


## Prep data
***

```{r read_input}

samples <- list.files(path = "../data/HGSC", pattern = "filtered_feature", recursive = TRUE, full.names = TRUE)
imgs <- list.files(path = "../data/HGSC", pattern = "hires_image", recursive = TRUE, full.names = TRUE)
spotfiles <- list.files(path = "../data/HGSC", pattern = "tissue_positions", recursive = TRUE, full.names = TRUE)
json <- list.files(path = "../data/HGSC", pattern = "scalefactors", recursive = TRUE, full.names = TRUE)

infoTable <- data.frame(samples, imgs, spotfiles, json, 
                        sample_id = paste0("sample_", 1:4),
                        site = c("HGSC_section1", "HGSC_section2", "HGSC_section3", "HGSC_section4"),
                        labels = c("site_1", "site_1", "site_2", "site_2"),
                        stringsAsFactors = FALSE)

se.ID103 <- InputFromTable(infoTable)

```

## Quality control
***

The average number of unique genes per spot is around 1k but overall the data is very sparse.

```{r qc}

QC <- function(object) {
  smpls <- GetStaffli(object)@meta.data$sample
  unique.smpls <- unique(smpls)
  df <- do.call(rbind, lapply(unique.smpls, function(s) {
    m <- object@assays$RNA@counts[, smpls == s]
    m <- data.frame(nGene = colSums(m > 0), nUMI = colSums(m), sample = s, stringsAsFactors = FALSE)
  }))
  df %>% group_by(sample) %>%
    summarize(mean_nGene = mean(nGene), median_nGene = median(nGene), max_nGene = max(nGene), min_nGene = min(nGene), 
              mean_nUMI = mean(nUMI), median_nUMI = median(nUMI), max_nUMI = max(nUMI), min_nUMI = min(nUMI))
}

qc.df <- QC(se.ID103)
qc.df

```

## Filter
***

Remove mitochondrial protein coding genes and enrich for protein coding genes


```{r filter, fig.width=14, fig.height=6}

gene_attr.ID103 <- data.frame(nUMI = Matrix::rowSums(se.ID103@assays$RNA@counts))

# Reload data with filters
se.ID103 <- InputFromTable(infoTable, min.gene.count = 10, min.spot.feature.count = 150)
# Remove mitochondrial genes
ensids <- read.table(file = "../data/annotation/GRCh38_genes.tsv", header = T, sep = "\t", stringsAsFactors = FALSE)
bio.keep <- c("protein_coding", "TR_V_gene", "TR_D_gene", "TR_J_gene", "TR_C_gene", "IG_LV_gene", "IG_V_gene", "IG_J_gene", "IG_C_gene", "IG_D_gene")
ensids <- subset(ensids, gene_biotype %in% bio.keep)
mt.genes <- grep(pattern = "^MT-", x = rownames(se.ID103), value = TRUE)

keep.genes <- setdiff(ensids$gene_name, mt.genes)
se.ID103 <- se.ID103[intersect(rownames(se.ID103), keep.genes), ]

```

## Normalize
***

```{r read_data, include=FALSE}
se.ID103 <- readRDS("../R_objects/se.ID103")
```
```{r normalize, eval=FALSE}

se.ID103 <- SCTransform(se.ID103, return.only.var.genes = FALSE, variable.features.n = NULL, variable.features.rv.th = 1.1)

```

## Load images
***

Section 1 and 2 are oriented quite differently so we'll align section 2 to section 1 using `ManualAlignImages`.

```{r load_images, eval=FALSE}

se.ID103 <- LoadImages(se.ID103, time.resolve = FALSE)
se.ID103 <- ManualAlignImages(se.ID103, type = "raw", edges = FALSE)

```


## NNMF
***

Run non-negative matrix factorization.

```{r NMF, eval=FALSE}

se.ID103 <- RunNMF(se.ID103, nfactors = 15)

```

### Functional Enrichment Analysis (FEA)

Here we select the top contributing genes for each factor using a simple mean + N*SD threshold. 
Top contributing genes are selected from the feature loadings of the NMF dimensionality reduction object. 
For each feature loading vector `vals`, the values are first log transformed, then any non-finite values are 
removed to produce a new vector `logvals`. 
Thereafter, genes with a value higher than mean(logvals) + 1.645*sd(logvals) are defined as top contributing genes.

```{r nmf_pathways}

# inflexion point selection
top.genes <- lapply(1:13, function(i) {
  vals <- se.ID103@reductions$NMF@feature.loadings[, i]
  logvals <- log(vals)
  logvals <- logvals[is.finite(logvals)]
  thr <- mean(logvals) + 1.645*sd(logvals)
  names(logvals[logvals > thr])
})

```

FEA is conducted using gprofiler2 and the Gene Ontology : Biological Processes (GO:BP) database as well 
as the cancer hallmark pathways from MSigDB.

```{r FEA}

pathways <- lapply(seq_along(top.genes), function(i) {
  gset <- top.genes[[i]]
  df1 <- gost(query = gset, organism = "mmusculus", sources = "GO:BP", evcodes = TRUE)$result
  df2 <- gost(query = gset, organism = 'gp__aJd5_EBf2_fFc', evcodes = TRUE)$result
  df <- rbind(df1, df2)
  if (is.null(df)) return(NULL)
  df$factor <- paste0("factor_", i)
  return(df)
})
pathways <- do.call(rbind, pathways)

pathways$factor <- factor(pathways$factor, paste0("factor_", 1:13))
pathways.summarized <- pathways %>% 
  group_by(factor) %>%
  top_n(n = 5, wt = -log10(p_value)) %>%
  mutate(GeneRatio = intersection_size/query_size) %>%
  arrange(factor(factor, levels = paste0("factor_", 13:1)), -p_value) %>%
  ungroup() %>%
  mutate(ord = 1:n()) 

sm <- pathways.summarized %>% group_by(factor) %>% summarize(x_start = min(ord)) %>% as.data.frame()
rownames(sm) <- sm$factor
sm <- sm[paste0("factor_", 13:1), ]
sm$x_end <- sm$x_start + c(diff(sm$x_start), nrow(pathways.summarized) - max(sm$x_start) + 1)
sm$x_start <- sm$x_start - 0.5
sm$x_end <- sm$x_end - 0.5
sm$x_mid <- (sm$x_end + sm$x_start)/2

# Remove term prefix
pathways.summarized$term_name[pathways.summarized$source == "h.all.v7.1.symbols"] <- pathways.summarized$term_id[pathways.summarized$source == "h.all.v7.1.symbols"]
pathways.summarized$term_name <- gsub(pattern = "_", replacement = " ", x = pathways.summarized$term_name)
pathways.summarized$term_name <- tolower(pathways.summarized$term_name)
pathways.summarized$face <- ifelse(pathways.summarized$source == "GO:BP", "plain", "bold")

```

```{r plot_pathways_results, fig.width=6, fig.height=10}
ggplot() +
  geom_rect(data = sm, aes(xmin = x_start, xmax = x_end, ymin = 0, ymax = 30, fill = factor), alpha = 0.5, color = "black") +
  geom_segment(data = pathways.summarized, aes(x = ord, xend = ord, y = 0, yend = -log10(p_value)), stat = "identity", linetype = "dashed") +
  geom_point(data = pathways.summarized, aes(ord, -log10(p_value), size = GeneRatio), stat = "identity", shape = 21, fill = "grey") +
  scale_x_continuous(position = "top", breaks = pathways.summarized$ord, labels = pathways.summarized$term_name) +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0, 30, length.out = 9), expand = c(0, 0)) +
  scale_size_continuous(range = c(0.5, 5)) +
  #guides(fill = FALSE) +
  coord_flip() +
  theme(legend.position = "bottom", 
        axis.text.y = element_text(face = pathways.summarized$face, size = 7), 
        plot.margin = margin(t = 1, r = 1, b = 0, l = 10), 
        axis.text.x = element_text(), 
        panel.background = element_blank(), 
        panel.grid.major.y = element_line(color = "lightgray"), 
        panel.grid.minor.y = element_line(color = "lightgray"), 
        plot.background = element_blank()) +
  labs(fill = "", x = "")

```

## Code for suppl. fig. 12
***

```{r violin_plots}

gg <- data.frame(unique_genes = se.ID103$nFeature_RNA, UMIs = se.ID103$nCount_RNA, site = se.ID103$site, sample = se.ID103$label)

p1 <- ggplot(gg, aes(sample, fill = site)) +
  geom_violin(aes(sample, unique_genes)) +
  geom_boxplot(aes(sample, unique_genes), width = 0.1) +
  labs(x = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
p2 <- ggplot(gg, aes(sample, fill = site)) +
  geom_violin(aes(sample, UMIs)) +
  geom_boxplot(aes(sample, UMIs), width = 0.1) +
  labs(x = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))

p.vln <- cowplot::plot_grid(p1, p2)
p.vln

```

```{r}

umis1 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section1")]
umis2 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section2")]
gene_attr_siteA <- data.frame(s1 = log10(rowSums(umis1) + 1), s2 = log10(rowSums(umis2) + 1))

umis1 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section3")]
umis2 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section4")]
gene_attr_siteB <- data.frame(s1 = log10(rowSums(umis1) + 1), s2 = log10(rowSums(umis2) + 1))

umis1 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section1", "HGSC_section2")]
umis2 <- se.ID103@assays$RNA@counts[, se.ID103$label %in% c("HGSC_section3", "HGSC_section4")]
gene_attr_siteAvB <- data.frame(s1 = log10(rowSums(umis1) + 1), s2 = log10(rowSums(umis2) + 1))

```

```{r fig.width=12, fig.height=5}
p1 <- ggplot() +
  geom_point(data = gene_attr_siteA, aes(s1, s2)) +
  geom_smooth(data = gene_attr_siteA, aes(s1, s2), method = lm) +
  labs(x = "log10(section_1 counts + 1)", y = "log10(section_2 counts + 1)", title = "") +
  ggpubr::stat_cor(data = gene_attr_siteA, aes(s1, s2)) +
  theme_minimal()

p2 <- ggplot() +
  geom_point(data = gene_attr_siteB, aes(s1, s2)) +
  geom_smooth(data = gene_attr_siteB, aes(s1, s2), method = lm) +
  labs(x = "log10(section_3 counts + 1)", y = "log10(section_4 counts + 1)", title = "") +
  ggpubr::stat_cor(data = gene_attr_siteB, aes(s1, s2)) +
  theme_minimal()

p3 <- ggplot() +
  geom_point(data = gene_attr_siteAvB, aes(s1, s2)) +
  geom_smooth(data = gene_attr_siteAvB, aes(s1, s2), method = lm) +
  labs(x = "log10(site_1 counts + 1)", y = "log10(site_2 counts + 1)", title = "") +
  ggpubr::stat_cor(data = gene_attr_siteAvB, aes(s1, s2)) +
  theme_minimal()

p.cor <- cowplot::plot_grid(p1, p2, p3, ncol = 3)
p.cor

```

Combine plots 

```{r, fig.width=12, fig.height=8}
p <- cowplot::plot_grid(p.vln, p.cor, ncol = 1) 
p
```

## Code for suppl. fig. 13
***

Spatial distribution

```{r, fig.width=18, fig.height=12}
ST.DimPlot(se.ID103, reduction = "NMF", center.zero = FALSE, dims = 1:5, grid.ncol = 5, cols = viridis::magma(n = 9) %>% rev(), show.sb = F)
```

Gene contribution

```{r, fig.width=12, fig.height=6}
plots <- lapply(1:5, function(f) {
  FactorGeneLoadingPlot(se.ID103, factor = f, topn = 10)
})
cowplot::plot_grid(plotlist = plots, ncol = 5)
```


## Date
***

```{r date}
date()
```

## Session info
***

```{r session}
sessionInfo()

```
