---
title: "Organoids"
author: "Ludvig Larsson"
date: "8/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(STutility)
library(magrittr)
library(dplyr)
library(Seurat)
```

# Prep data
***

Prepare spaceranger output and load into a Seurat object.

```{r load_data}
samples <- list.files("../data/kidneyorganoid", pattern = "filtered_feature_bc_matrix.h5", recursive = T, full.names = T)
imgs <- list.files("../data/kidneyorganoid", pattern = "tissue_hires_image.png", recursive = T, full.names = T)
spotfiles <- list.files("../data/kidneyorganoid", pattern = "tissue_positions_list.csv", recursive = T, full.names = T)
json <- list.files("../data/kidneyorganoid", pattern = "scalefactors_json.json", recursive = T, full.names = T)

infoTable <- data.frame(samples, imgs, spotfiles, json,
                        sample_id = paste0("sample_", 1:3), 
                        stringsAsFactors = F)

se <- InputFromTable(infoTable)

```

```{r filter_genes}

ensids <- read.table(file = "~/Eduardo/misc/GRCh38_genes.tsv", header = T, stringsAsFactors = F)
rownames(ensids) <- ensids$gene_name
keep.biotypes <- subset(ensids, gene_biotype %in% "protein_coding")$gene_name

# Protein coding genes to remove
rm.genes <- grep(pattern = "^RPL|^RPS|^MT-", x = rownames(se), value = T)

# Determine genes to keep
keep.genes <- setdiff(intersect(rownames(se), keep.biotypes), rm.genes)

# Filter data
se.filtered <- se[keep.genes, ]

```

## Load images
***

```{r}

se.filtered <- LoadImages(se.filtered, time.resolve = FALSE)

```

## Find isolated islets
***

This was done by manual labelling of spots from each organoid. Manual labelling can be done with
STUtility using `ManualAnnotation()`.

```{r selection, eval=FALSE, fig.width=8, fig.height=8}

organoid <- read.table("../data/kidneyorganoid/organoid_selection.txt")
se.filtered$organoid <- organoid[rownames(se.filtered@meta.data), "organoid"]
#se.filtered <- ManualAnnotation(se.filtered)
#se.filtered$organoid <- se.filtered$labels

ST.FeaturePlot(se.filtered, features = "organoid", ncol = 2)

```

## Crop data
***

Here we can use the `CropImages()` function to crop out each organoid into a new dataset. 

Note that if you want to work with higher resolution cropped images, you will need to use higher 
resolution versions of the HE images. The raw HE images are not available in this repo. 

```{r crop}

library(magick)
# Subset data
se.subset.filtered <- SubsetSTData(se.filtered, spots = rownames(subset(se.filtered[[]], organoid %in% paste0("Org", c(1, 3:10)))))

# Get crop windows
pxs <- cbind(GetStaffli(se.subset.filtered)@meta.data[, c("pixel_x", "pixel_y", "sample")], selection = se.subset.filtered$organoid)
pxs.split <- split(pxs, pxs$sample)
orgs <- c()
pxs.split <- setNames(lapply(pxs.split, function(pxs) {
  sel.split <- lapply(split(pxs, pxs$selection), function(pxs) {
    orgs <<- c(orgs, unique(pxs$selection))
    m <- apply(pxs[, 1:2], 2, range)
    centroid <- apply(m , 2, mean)
    widths <- apply(m, 2, function(x) round(diff(range(x)))) + 400
    offsets <- centroid - max(widths)/2
    geometry <- geometry_area(width = max(widths), height = max(widths), x_off = max(offsets[1], 0), y_off = max(offsets[2], 0))
  })
}), nm = unique(pxs$sample))

# Collect crop areas
crop.windows <- do.call(c, lapply(unique(pxs$sample), function(s) {
  crop_geometries <- setNames(pxs.split[[s]], nm = rep(s, length(pxs.split[[s]])))
}))

# Crop data
organoids <- paste0("Org", c(1, 3:10))
se.cropped.list <- lapply(organoids, function(org) {
  crp <- crop.windows[orgs == org]
  se <- CropImages(se.subset.filtered, crop.geometry.list = crp, time.resolve = TRUE, verbose = TRUE)
  return(se)
})

# Filter out spots that are not matching organoid id
se.cropped.list <- lapply(seq_along(se.cropped.list), function(i) {
  SubsetSTData(se.cropped.list[[i]], expression = organoid %in% organoids[i])
})
merged.data <- MergeSTData(x = se.cropped.list[[1]], y = se.cropped.list[2:length(se.cropped.list)])

```

Subset organoids

```{r subset}

subset.merged.data <- SubsetSTData(merged.data, expression = organoid %in% paste0("Org", c(1:4, 8:10)))

```

```{r, fig.width=8, fig.height=8}

ImagePlot(subset.merged.data, method = "raster")

```


Analysis

```{r analysis}

subset.merged.data <- subset.merged.data %>% 
  SCTransform(return.only.var.genes = FALSE) %>%
  RunPCA() %>%
  RunUMAP(dims = 1:20) %>%
  FindNeighbors(reduction = "pca", dims = 1:20) %>%
  FindClusters(resolution = 0.3)

```

# Count QC metrics per organoid

```{r qc_per_organoid}

sapply(paste0("Org", c(1, 3:4, 8:10)), function(org) {
  mean(subset(subset.merged.data[[]], organoid %in% org)$nFeature_RNA)
})

```

Plots for suppl. fig. 11

```{r, fig.width=8, fig.height=4}

subset.merged.data$organoid <- factor(subset.merged.data$organoid, levels = paste0("Org", c(1:4, 8:10)))
p1 <- DimPlot(subset.merged.data, group.by = "organoid", cols = c("#332288", "#88CCEE", "#117733", "#DDCC77", "#CC6677","#AA4499"), pt.size = 1.3)
p2 <- DimPlot(subset.merged.data, group.by = "seurat_clusters", pt.size = 1.3)
p1 - p2

```

```{r, fig.width=8, fig.height=8}

de.markers <- FindAllMarkers(subset.merged.data)
top10 <- de.markers %>%
  group_by(cluster) %>%
  filter(avg_log2FC > 0, p_val_adj < 0.01) %>%
  top_n(wt = -p_val_adj, n = 10)
p <- DoHeatmap(subset.merged.data, features = top10$gene)
p

```

QC features

```{r}

VlnPlot(subset.merged.data, features = "nFeature_RNA", group.by = "organoid")

```

## Date
***

```{r date}
date()
```

## Session
***

```{r session}
sessionInfo()
```

