---
title: "Analysis of FFPE mouse brain"
author: "Ludvig Larsson"
date: "5/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)

```

# Load data
***

Prepare spaeranger output files and load data into Seurat object.

```{r read_input}

samples <- list.files("../data/lung_SARS-CoV-2", pattern = "filtered", recursive = T, full.names = T)
imgs <- list.files("../data/lung_SARS-CoV-2", pattern = "small", recursive = T, full.names = T)
spotfiles <- list.files("../data/lung_SARS-CoV-2", pattern = "tissue_positions", recursive = T, full.names = T)
json <- list.files("../data/lung_SARS-CoV-2", pattern = "scalefactors", recursive = T, full.names = T)

infoTable <- data.frame(samples, imgs, spotfiles, json,
                        sample_id = c("sample_1", "sample_2"), 
                        stringsAsFactors = F)

se <- InputFromTable(infoTable, scaleVisium = 0.3)

```

## Gene filter
***

Here I have enriched all datasets for gene labelled with a biotype of one of the following; "protein_coding", "TR_V_gene", "TR_D_gene", "TR_J_gene", "TR_C_gene", "IG_LV_gene", "IG_V_gene", "IG_J_gene", "IG_C_gene", "IG_D_gene"

This will filter out all "ncRNA" och keep only protein coding genes. I also removed all protein coding genes located in the mitochondrial genome. Then, as a final filter on spot level I excluded all spots containg 100 genes or less (after filtering out the previously mentioned genes).

```{r filter_data}

ensids <- read.table("../data/annotation/GRCh38_genes.tsv", header = TRUE, stringsAsFactors = FALSE)
ensids$gene_name <- paste0("GRCh38--", ensids$gene_name)
rownames(ensids) <- ensids$gene_name

gene_attr <- data.frame(gene = rownames(se), biotype = ensids[rownames(se), ]$gene_biotype, count = Matrix::rowSums(se@assays$RNA@counts))
gene_attr.summarized <- gene_attr %>% 
  dplyr::group_by(biotype) %>%
  dplyr::summarize(count = sum(count)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(count = count/sum(count))

ggplot(gene_attr.summarized, aes(x = "", y = count, fill = biotype)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0)

# Define what biotypes to keep
bio.keep <- c("protein_coding", "TR_V_gene", "TR_D_gene", "TR_J_gene", "TR_C_gene", "IG_LV_gene", "IG_V_gene", "IG_J_gene", "IG_C_gene", "IG_D_gene")
genes.keep <- intersect(rownames(se), subset(ensids, gene_biotype %in% bio.keep)$gene_name)

# Define what genes to remove
genes.remove <- grep(pattern = "GRCh38--MT-", x = rownames(se), value = TRUE)

# Define what genes to keep
genes.keep <- setdiff(genes.keep, genes.remove)
genes.keep <- c(genes.keep, "wuhCor1-ORF1ab")

# Filter data
se <- se[genes.keep, ]
se$nFeature_RNA <- Matrix::colSums(se@assays$RNA@counts > 0)
se$nCount_RNA <- Matrix::colSums(se@assays$RNA@counts)
se <- SubsetSTData(se, spots = rownames(subset(se[[]], nFeature_RNA > 100)))

rownames(se@assays$RNA@counts) <- gsub(pattern = "GRCh38--", replacement = "", x = rownames(se))
rownames(se@assays$RNA@data) <- gsub(pattern = "GRCh38--", replacement = "", x = rownames(se))

```

## Quality Control
***

```{r QC}

se <- SetIdent(se, value = "sample_id")
gg <- data.frame(sample = se$sample_id, 
                 unique_genes = se$nFeature_RNA,
                 UMIs = se$nCount_RNA)

p1 <- ggplot(gg, aes(sample, unique_genes, fill = sample)) +
  geom_violin() +
  geom_boxplot(width = 0.1) +
  theme_minimal()
p2 <- ggplot(gg, aes(sample, UMIs, fill = sample)) +
  geom_violin() +
  geom_boxplot(width = 0.1) +
  theme_minimal()
p1 - p2

```

## Analysis of FFPE section
***

```{r run_analysis, fig.width=9, fig.height=7}

se <- se %>%
  SCTransform(return.only.var.genes = FALSE)

se <- se %>% RunNMF(nfactors = 15, n.cores = 7)

se <- se %>%
  #RunPCA() %>%
  FindNeighbors(dims = 1:15, reduction = "NMF") %>%
  FindClusters(resolution = 0.6) %>%
  RunUMAP(dims = 1:15, reduction = "NMF", n.epochs = 1e3)
  
```

factor heatmap

```{r, fig.width=4, fig.height=9}

library(zeallot)
c(df1, df2) %<-% SummarizeAssocFeatures(se, features.return = 6)
p <- swne::ggHeat(df2[nrow(df2):1, ], rescaling = "column") +
  scale_fill_gradientn(colors = viridis::magma(n = 9) %>% rev())
p

```


Crop data

```{r}

se <- LoadImages(se, time.resolve = F)
crop.windows <- list(TLS = c("1" = "2843x2482+7438+1595", "2" = "2402x2578+4235+5925"),
                     SM = c("1" = "3382x3066+6730+5908", "2" = "3149x3348+8435+5700"))

se.cropped.list <- lapply(crop.windows, function(crop.w) {
  CropImages(se, crop.geometry.list = crop.w, xdim = 2e3)
})

```

Display images

```{r}
ImagePlot(se.cropped.list[[1]], method = "raster")
ImagePlot(se.cropped.list[[2]], method = "raster")
```


Example gene

```{r, fig.width=10, fig.height=5}

FeatureOverlay(se.cropped.list[[1]], features = "PIGR", sampleids = 1:2, ncols.features = 1,
               ncols = 2, show.sb = F, pt.size = 8, value.scale = "all")

```

Main figure 4

```{r fig.width=12, fig.height=4}

i <- 1
max_val <- max(max(se.cropped.list[[1]]@reductions$NMF@cell.embeddings[, paste0("factor_", i)]),
           max(se.cropped.list[[2]]@reductions$NMF@cell.embeddings[, paste0("factor_", i)]))
p1 <- FeatureOverlay(se.cropped.list[[1]], features = paste0("factor_", i), sample.label = F,
                     max.cutoff = "q99", pt.size = 4.5,
                     value.scale = c(0, max_val),
                     cols = c("lightgray", "yellow"),
                     add.alpha = T, show.sb = FALSE)
p2 <- FeatureOverlay(se.cropped.list[[2]], features = paste0("factor_", i), sample.label = F,
                     max.cutoff = "q99", pt.size = 4,
                     value.scale = c(0, max_val),
                     cols = c("lightgray", "yellow"),
                     add.alpha = T, show.sb = FALSE)
p3 <- FactorGeneLoadingPlot(se.cropped.list[[1]], factor = i, topn = 20)

cowplot::plot_grid(p1, p2, p3, ncol = 3, rel_widths = c(1, 1, 1))

```



Single cell markers

```{r}

library(readxl)
path <- "../data/lungcellmarkers/41586_2020_2922_MOESM6_ESM.xlsx"
sheetnames <- excel_sheets(path)
sheetnames <- sheetnames[-grep(pattern = "SS", x = sheetnames)]
nms <- c()
mylist <- lapply(sheetnames, function(xlp) {
  xl <- read_excel(path = path, sheet = xlp)
  nms <<- c(nms, colnames(xl)[1])
  xl <- read_excel(path = path, sheet = xlp, skip = 1)
})
mylist <- setNames(mylist, nm = nms)

GeneSets <- lapply(mylist, function(x) {
  (x %>% filter(avg_logFC > 0.8, p_val_adj < 0.01))$Gene
})

```


Cor AUC with factors

```{r fig.width=10, fig.height=8}

library(AUCell)

spots_rankings <- AUCell_buildRankings(se@assays$SCT@data, nCores = 7, plotStats = FALSE)
cells_AUC <- AUCell_calcAUC(GeneSets, spots_rankings, aucMaxRank = ceiling(0.05 * nrow(spots_rankings)))

NMF <- se@reductions$NMF@cell.embeddings
se[["AUC"]] <- CreateAssayObject(data = getAUC(cells_AUC))
AUC <- se[["AUC"]]@data
AUC <- AUC[c("B", "Plasma", "CD4+ Naive T", "CD8+ Naive T", 
             "CD8+ Memory/Effector T", "CD4+ Memory/Effector T",
             "Ciliated", "Basal", "Serous", "Goblet", "Mucous",
             "Alveolar Epithelial Type 1", "Alveolar Epithelial Type 2",
             "Lymphatic", "Airway Smooth Muscle", "Vascular Smooth Muscle",
             "Macrophage", "Alveolar Fibroblast", "Fibromyocyte", "Myofibroblast" ,
              "Classical Monocyte", "Capillary", "Myeloid Dendritic Type 1",
             "Myeloid Dendritic Type 2",
              "Nonclassical Monocyte"), ]
cors <- setNames(as.data.frame(sapply(1:15, function(i){
  setNames(unlist(lapply(rownames(AUC), function(n) {
    cor(log1p(NMF[, i]), log1p(AUC[n, ]))
  })), rownames(AUC))
})), paste0("factor_", 1:15))

m <- t(cors)
#p <- pheatmap::pheatmap(m, border_color = "white", 
#                   color = colorRampPalette(RColorBrewer::brewer.pal(n = 11, name = "RdBu") %>% rev())(50),
#                   breaks = seq(-max(m), max(m), length.out = 51))

d.row <- dist(m)
tree.row <- hclust(d.row, method = "ward.D2")
d.col <- dist(t(m))
tree.col <- hclust(d.col, method = "ward.D2")

mm <- reshape2::melt(m)
mm$Var1 <- factor(mm$Var1, levels = rownames(m)[tree.row$order])
mm$Var2 <- factor(mm$Var2, levels = colnames(m)[rev(tree.col$order)])

p <- ggplot(mm) +
  geom_tile(aes(Var1, Var2), fill = NA, color = "lightgray") +
  geom_point(aes(Var1, Var2, size = abs(value), color = value)) +
  guides(size = F) +
  scale_size_continuous(range = c(2, 7)) +
  scale_color_gradientn(colours = RColorBrewer::brewer.pal(n = 11, name = "RdBu") %>% rev(), 
                        limits = c(-max(mm$value), max(mm$value)),
                        breaks = c(-0.6, -0.3, 0, 0.3, 0.6)) +
  theme_minimal() +
  labs(x = "",  y = "", color = "correlation\nscore") +
  theme(panel.grid = element_blank(), 
        axis.text.x = element_text(angle = 30, hjust = 1, color = "black"),
        axis.text.y = element_text(color = "black"))

p

```

## Date 
***

```{r}
date()
```

## Session
***

```{r}
sessionInfo()
```


