---
title: "Organoids"
author: "Ludvig Larsson"
date: "8/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(STutility)
library(magrittr)
library(dplyr)
library(Seurat)
```

## Prep data
***

Prepare spaceranger output files and load data into a Seurat object.

```{r load_data}
samples <- list.files("../data/lungorganoid", pattern = "filtered", recursive = T, full.names = T)
imgs <- list.files("../data/lungorganoid", pattern = "tissue_hires_image.png", recursive = T, full.names = T)
spotfiles <- list.files("../data/lungorganoid", pattern = "tissue_positions_list.csv", recursive = T, full.names = T)
json <- list.files("../data/lungorganoid", pattern = "scalefactors_json.json", recursive = T, full.names = T)

infoTable <- data.frame(samples, imgs, spotfiles, json,
                        sample_id = c("A1", "B1", "C1", "D1"), 
                        stringsAsFactors = F)

se <- InputFromTable(infoTable)

```

Enrich data fro protein coding genes.

```{r filter_genes}

ensids <- read.table(file = "../data/annotation/GRCh38_genes.tsv", header = T, stringsAsFactors = F)
rownames(ensids) <- ensids$gene_name
keep.biotypes <- subset(ensids, gene_biotype %in% "protein_coding")$gene_name

# Protein coding genes to remove
rm.genes <- grep(pattern = "^RPL|^RPS|^MT-", x = rownames(se), value = T)

# Determine genes to keep
keep.genes <- setdiff(intersect(rownames(se), keep.biotypes), rm.genes)

# Filter data
se.filtered <- se[keep.genes, ]

```

## Load images

```{r load images}

se.filtered <- LoadImages(se.filtered, time.resolve = FALSE)

```

## Annotate isolated islets

This was done by manual labelling of spots from each organoid. Manual labelling can be done with
STUtility using `ManualAnnotation()`.

```{r selection, eval=FALSE, fig.width=8, fig.height=8}

organoid <- read.table("../data/lungorganoid/organoid_selection.txt")
se.filtered$organoid <- organoid[rownames(se.filtered@meta.data), "organoid"]
#se.filtered <- ManualAnnotation(se.filtered)
#se.filtered$organoid <- se.filtered$labels

ST.FeaturePlot(se.filtered, features = "organoid", ncol = 2)

```

## Crop data

```{r crop}

library(magick)
# Subset data
se.subset.filtered <- SubsetSTData(se.filtered, spots = rownames(subset(se.filtered[[]], organoid %in% paste0("Org", 1:6))))

# Get crop windows
pxs <- cbind(GetStaffli(se.subset.filtered)@meta.data[, c("pixel_x", "pixel_y", "sample")], selection = se.subset.filtered$organoid)
pxs.split <- split(pxs, pxs$sample)
pxs.split <- setNames(lapply(pxs.split, function(pxs) {
  sel.split <- lapply(split(pxs, pxs$selection), function(pxs) {
    m <- apply(pxs[, 1:2], 2, range)
    centroid <- apply(m , 2, mean)
    widths <- apply(m, 2, function(x) round(diff(range(x)))) + 20 # Increase if you are using higher res HE images
    offsets <- centroid - max(widths)/2
    geometry <- geometry_area(width = max(widths), height = max(widths), x_off = pmax(offsets[1], 0), y_off = pmax(offsets[2], 0))
  })
}), nm = unique(pxs$sample))

# Collect crop areas
crop.windows <- do.call(c, lapply(unique(pxs$sample), function(s) {
  crop_geometries <- setNames(pxs.split[[s]], nm = rep(s, length(pxs.split[[s]])))
}))

# Crop data
se.cropped <- CropImages(se.subset.filtered, crop.geometry.list = crop.windows, time.resolve = TRUE, verbose = TRUE)

```


```{r plot_data, fig.height=8, fig.width=8}

ImagePlot(se.cropped, method = "raster")

```

## Run analysis
***

```{r analysis}

se.cropped <- se.cropped %>% 
  SCTransform(return.only.var.genes = FALSE) %>%
  RunPCA() %>%
  RunUMAP(dims = 1:20) %>%
  FindNeighbors(reduction = "pca", dims = 1:20) %>%
  FindClusters() %>% 
  FindClusters(resolution = 0.3)

```

Plot clusters on each organoid

```{r, fig.width=6.8, fig.height=10}

s <- data.frame(organoid = se.cropped$organoid, sample = GetStaffli(se.cropped)@meta.data$sample) 
s <- s[!duplicated(s), ] %>%
  arrange(organoid)

p.list <- lapply(as.integer(s$sample), function(i) {
  FeatureOverlay(se.cropped, features = "seurat_clusters", 
               sampleids = i,
               sample.label = F, show.sb = F,
               custom.theme = theme(
                 legend.position = "none", 
                 plot.margin = margin(0.1,0.1,0.1,0.1, "cm"),
                 plot.title = element_blank()
               )) 
})
cowplot::plot_grid(plotlist = p.list, ncol = 4)

```

QC features

```{r}

se.cropped <- SetIdent(se.cropped, value = "organoid")
VlnPlot(se.cropped, features = "nFeature_RNA", group.by = "organoid")

```


```{r, fig.width=8, fig.height=4}

p1 <- DimPlot(se.cropped, group.by = "organoid", cols = c("#332288", "#88CCEE", "#117733", "#DDCC77", "#CC6677","#AA4499"))
p2  <- DimPlot(se.cropped, group.by = "seurat_clusters")
p1 - p2

```

```{r, fig.width=8, fig.height=8}

de.markers <- FindAllMarkers(se.cropped)
top10 <- de.markers %>%
  group_by(cluster) %>%
  filter(avg_log2FC > 0) %>%
  top_n(wt = -p_val_adj, n = 10)
p <- DoHeatmap(se.cropped, features = top10$gene)
p

```

## Date
***

```{r}
date()
```


## Session
***

```{r}
sessionInfo()
```


