---
title: "Analysis of FFPE mouse brain"
author: "Ludvig Larsson"
date: "5/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

NOTE: Make sure that you have sctransform v0.2.1 installed to reproduce the plots. Later versions might change the normalization procedure and affect the outcome.

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)
library(harmony)

```

# Load data
***

```{r read_input}

samples <- c("../data/mousebrain/V19T26-039_A1/filtered_feature_bc_matrix.h5",
             "../data/mousebrain/FF1_public_data/V1_Adult_Mouse_Brain_filtered_feature_bc_matrix.h5")

imgs <- c("../data/mousebrain/V19T26-039_A1/spatial/tissue_hires_image.png",
          "../data/mousebrain/FF1_public_data/spatial/tissue_hires_image.png")

spotfiles <- c("../data/mousebrain/V19T26-039_A1/spatial/tissue_positions_list.csv",
               "../data/mousebrain/FF1_public_data/spatial/tissue_positions_list.csv")

json <- c("../data/mousebrain/V19T26-039_A1/spatial/scalefactors_json.json",
          "../data/mousebrain/FF1_public_data/spatial/scalefactors_json.json")

infoTable <- data.frame(samples, imgs, spotfiles, json, condition = c("FFPE", "FF"), sample_id = c("sample_1", "sample_2"), stringsAsFactors = F)

se <- InputFromTable(infoTable)

```

## Gene filter
***

Here I have enriched all datasets for gene labelled with a biotype of one of the following: 
"protein_coding", "TR_V_gene", "TR_D_gene", "TR_J_gene", "TR_C_gene", "IG_LV_gene", "IG_V_gene", 
"IG_J_gene", "IG_C_gene", "IG_D_gene"

This will filter out all "ncRNA" och keep only protein coding genes. I also removed all protein 
coding genes located in the mitochondrial genome. Then, as a final filter on spot level I excluded 
all spots containg 100 genes or less (after filtering out the previously mentioned genes).

```{r filter_data}

ensids <- read.table("../data/annotation/mm10_genes.tsv", header = TRUE, stringsAsFactors = FALSE)
rownames(ensids) <- ensids$gene_name

# Define what biotypes to keep
bio.keep <- c("protein_coding", "TR_V_gene", "TR_D_gene", "TR_J_gene", "TR_C_gene", "IG_LV_gene", "IG_V_gene", "IG_J_gene", "IG_C_gene", "IG_D_gene")
genes.keep <- intersect(rownames(se), subset(ensids, gene_biotype %in% bio.keep)$gene_name)

# Define what genes to remove
genes.remove <- grep(pattern = "^mt-", x = rownames(se), value = TRUE)

# Define what genes to keep
genes.keep <- setdiff(genes.keep, genes.remove)

# Filter data
se <- se[genes.keep, ]
se$nFeature_RNA <- Matrix::colSums(se@assays$RNA@counts > 0)
se$nCount_RNA <- Matrix::colSums(se@assays$RNA@counts)
se <- SubsetSTData(se, spots = rownames(subset(se[[]], nFeature_RNA > 100)))

```

## Quality Control
***

Distribution of unqiue genes and UMIs per spot.

```{r QC}

se <- SetIdent(se, value = "condition")
VlnPlot(se, features = c("nFeature_RNA", "nCount_RNA"), log = T, pt.size = 0.5)

```

### Load and process images
***

Here we'll apply a mask rotate the FFPE tissue section 

```{r load_images_all}

se <- LoadImages(se, time.resolve = FALSE)
se <- MaskImages(se)
se <- WarpImages(se, transforms = list("1" = list(angle = 60)))

```

## Analysis of FFPE section
***

```{r run_analysis, eval=FALSE}

se.FFPE <- SubsetSTData(se, spots = rownames(subset(se[[]], condition %in% "FFPE")))

se.FFPE <- se.FFPE %>%
  SCTransform(variable.features.rv.th = 1.1, variable.features.n = NULL, return.only.var.genes = FALSE) %>%
  RunPCA() %>%
  RunUMAP(dims = 1:25, reduction = "pca", n.epochs = 1e3, n.neighbors = 30) %>%
  FindNeighbors(dims = 1:25, reduction = "pca") %>% 
  FindClusters(resolution = 1.4)

```
```{r run_analysis, eval=FALSE}

se.FF <- SubsetSTData(se, spots = rownames(subset(se[[]], condition %in% "FF")))

se.FF <- se.FF %>%
  SCTransform(variable.features.rv.th = 1.1, variable.features.n = NULL, return.only.var.genes = FALSE) %>%
  RunPCA() %>%
  RunUMAP(dims = 1:25, reduction = "pca", n.epochs = 1e3, n.neighbors = 30) %>%
  FindNeighbors(dims = 1:25, reduction = "pca") %>% 
  FindClusters(resolution = 0.8)

```

## Load registrations

```{r load_regi}

# Load functions
source("global_functions.R")

# Load registration objects created with the wholebrain workflow
regi.FFPE <- readRDS("../data/registration/registration_HE_FFPE_mouse_brain")
regi.FF <- readRDS("../data/registration/registration_HE_FF_mouse_brain")

# Extract x, y coordinates
spots.FFPE <- setNames(GetStaffli(se.FFPE)@meta.data[, c("warped_x", "warped_y")], nm = c("x", "y"))
spots.FF <- setNames(GetStaffli(se.FF)@meta.data[, c("warped_x", "warped_y")], nm = c("x", "y"))

# Obtain atlas annaotations of spots
annotated.spots.FFPE <- get_allen_brain_annotations(regi.FFPE, xy = spots.FFPE)
annotated.spots.FF <- get_allen_brain_annotations(regi.FF, xy = spots.FF)

# Load the Allen Brain Atlas ontology
library(wholebrain)
data("atlasOntology")

# Annotate spots at a higher level (FFPE)
d.FFPE <- find_parents(na.omit(unique(annotated.spots.FFPE$acronym)), target_acronyms = c("STR", "CTXsp", "CA1sp", "CA3sp", "DG-sg", "VL", "HY", "fiber tracts", "TH", "OLF", "Isocortex"), ann = atlasOntology)
se.FFPE$acronym <- annotated.spots.FFPE[colnames(se.FFPE), "acronym"]
se.FFPE$target_acronym <- d.FFPE[se.FFPE$acronym, "target_acronym"]
se.FFPE$target_acronym <- factor(se.FFPE$target_acronym, levels = names(sort(table(se.FFPE$target_acronym), decreasing = T)))

# Annotate spots at a higher level (FF)
d.FF <- find_parents(na.omit(unique(annotated.spots.FF$acronym)), target_acronyms = c("STR", "CTXsp", "CA1sp", "CA3sp", "DG-sg", "VL", "HY", "fiber tracts", "TH", "OLF", "Isocortex"), ann = atlasOntology)
se.FF$acronym <- annotated.spots.FF[colnames(se.FF), "acronym"]
se.FF$target_acronym <- d.FF[se.FF$acronym, "target_acronym"]
se.FF$target_acronym <- factor(se.FF$target_acronym, levels = names(sort(table(se.FFPE$target_acronym), decreasing = T)))


```

### DE analysis of FFPE section clusters
***

```{r de_analysis, eval=FALSE}
# DE-analysis
se.FFPE <- SetIdent(se.FFPE, value = "target_acronym")
se.FFPE.subset <- SubsetSTData(se.FFPE, spots = names(se.FFPE$target_acronym[!is.na(se.FFPE$target_acronym)]))
de.markers.FFPE <- FindAllMarkers(se.FFPE.subset, only.pos = F)
top10.FFPE <- de.markers.FFPE %>%
  dplyr::group_by(cluster) %>%
  dplyr::filter(avg_logFC > 0) %>%
  dplyr::top_n(n = 7, wt = -log10(p_val_adj)) %>%
  dplyr::arrange(cluster, avg_logFC)
```
```{r de_analysis, eval=FALSE}

# DE-analysis
se.FF <- SetIdent(se.FF, value = "target_acronym")
se.FF.subset <- SubsetSTData(se.FF, spots = names(se.FF$target_acronym[!is.na(se.FF$target_acronym)]))
de.markers.FF <- FindAllMarkers(se.FF.subset, only.pos = F)
top10.FF <- de.markers.FF %>%
  dplyr::group_by(cluster) %>%
  dplyr::filter(avg_logFC > 0) %>%
  dplyr::top_n(n = 7, wt = -log10(p_val_adj)) %>%
  dplyr::arrange(cluster, avg_logFC)
```

## Check gene overlap

```{r geneoverlap, fig.width=5, fig.height=4}

lvls <- c("CA1sp", "CA3sp", "CTXsp", "DG-sg", "fiber tracts", "HY", "Isocortex", "OLF", "STR", "TH", "VL")
de.markers.FF.filtered <- de.markers.FF %>% filter(abs(avg_logFC) > 0.3)
de.markers.FFPE.filtered <- de.markers.FFPE %>% filter(abs(avg_logFC) > 0.3)

enriched.genes.FFPE <- pairwise.scores(se.FFPE)
enriched.genes.FF <- pairwise.scores(se.FF)

res <- sapply(lvls, function(lvl.A) {
  setNames(sapply(lvls, function(lvl.B) {
    listA = setNames(enriched.genes.FF[, lvl.A],
                     rownames(enriched.genes.FF))
    listA <- sort(listA, decreasing = T)[1:1000]
    listB = setNames(enriched.genes.FFPE[, lvl.B],
                     rownames(enriched.genes.FFPE))
    listB <- sort(listB, decreasing = T)[1:1000]
    rbo(listA, listB, p = 0.95)
  }), lvls)
})

cols <- c("Isocortex" = "#ABDDA4", "CTXsp" = "#66C2A5", "OLF" = "#3288BD",
          "CA3sp" = "#5E4FA2", "CA1sp" = "#ba8cd7", "DG-sg" = "#9E0142",
          "TH" = "#F46D43", "HY" = "#FDAE61", 
          "VL" = "#FEE08B", "STR" = "#FFFFBF", "fiber tracts" = "lightgray")
o <- c("Isocortex", "CTXsp", "OLF", "CA3sp", "CA1sp", "DG-sg", "TH", "HY", "VL", "STR", "fiber tracts")
ann <- data.frame("region" = o, row.names = o)
ann.cols <- list("region" = cols)
p <- pheatmap::pheatmap(res[o, o], cluster_cols = F, cluster_rows = F, border_color = NA, 
                   color = colorRampPalette(RColorBrewer::brewer.pal(n = 9, name = "Oranges"))(50), 
                   annotation_row = ann, annotation_col = ann, 
                   annotation_colors = ann.cols, labels_col = "", labels_row = "")
p

save_pheatmap_pdf <- function(x, filename, width = 7, height = 7) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width = width, height = height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}
#save_pheatmap_pdf(p, "../figures/figure_1/Overlap_heatmap.pdf", width = 10, height = 8)

```


## stereoscope
***

Load cell type proportions and store in Seurat object as a new assay.

```{r celltype_mapping_results, fig.width=20, fig.height=30}

props.FFPE <- data.frame(read.table(file = "../data/stereoscope/ffpe-prop.tsv", sep = "\t", header = T), row.names = 1)
rownames(props.FFPE) <- paste(rownames(props.FFPE), 1, sep = "_")

props.FF <- data.frame(read.table(file = "../data/stereoscope/ff-prop.tsv", sep = "\t", header = T), row.names = 1)
rownames(props.FF) <- paste(rownames(props.FF), 2, sep = "_")

all.props <- rbind(props.FFPE, props.FF)
all.props.subset <- t(all.props[colnames(se), ])

assay <- CreateAssayObject(data = all.props.subset)

se[["stereoscope"]] <- assay
DefaultAssay(se) <- "stereoscope"
ST.FeaturePlot(se, features = colnames(all.props), grid.ncol = 4, show.sb = FALSE)

```

Sum proportions over entire tissue section and display as bar charts.

```{r sum_props}

gg <- sapply(unique(se$condition), function(i) {
  x <- rowSums(all.props.subset[, colnames(se)[se$condition %in% i]])
  x/sum(x)
})
ggm <- reshape2::melt(gg)
ggm$value <- ifelse(ggm$Var2 == "FF", -ggm$value, ggm$value)
p <- ggplot(ggm, aes(x = Var1, value, fill = Var2)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = c(-0.15, 0, 0.15), labels = c(0.15, 0, 0.15)) +
  labs(x = "cell type", y = "proportion", fill = "tissue type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1, color = "black", size = 9))
p

```

## Date()
***

```{r date}

date()

```

## Session info
***

```{r session_info}

sessionInfo()

```
