---
title: "mouse brain multiple sections"
author: "Ludvig Larsson"
date: "12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

NOTE: Make sure that you have sctransform v0.2.1 installed to reproduce the plots. Later versions might change the normalization procedure.

```{r load_libs, warning=FALSE, message=FALSE}

library(STutility)
library(magrittr)
library(dplyr)
library(harmony)

```

# Load data
***

Here we have included 7 FFPE mouse brain section datasets for the analysis.

```{r read_input}

dirs <- c("../data/mousebrain/V10F24-078_A1",
          "../data/mousebrain/V10F24-078_B1", "../data/mousebrain/V10F24-078_C1",
          "../data/mousebrain/V10F24-078_D1", "../data/mousebrain/V10F24-110_A1",
          "../data/mousebrain/V10F24-110_B1", "../data/mousebrain/V19T26-039_A1")
samples <- sapply(dirs, function(p) {list.files(path = p, pattern = "filtered", recursive = T, full.names = T)})
imgs <- sapply(dirs, function(p) {list.files(path = p, pattern = "tissue_hires_image.png", recursive = T, full.names = T)})
spotfiles <- sapply(dirs, function(p) {list.files(path = p, pattern = "tissue_positions_list.csv", recursive = T, full.names = T)})
json <- sapply(dirs, function(p) {list.files(path = p, pattern = "scalefactors_json.json", recursive = T, full.names = T)})


infoTable <- data.frame(samples, imgs, spotfiles, json, 
                        condition = c( "normal", "normal", "normal", "normal",  "RU", "RU", "normal"),
                        date = c("200925", "200925", "200925", "200925", "201016", "201016", "200513"), 
                        sample_id = c("FFPE1", "FFPE2", "FFPE3", "FFPE4", "FFPE5", "FFPE6", "FFPE7"), 
                        stringsAsFactors = F)
infoTable <- infoTable[c(7, 1:6), ]

se <- InputFromTable(infoTable)

```

## Analysis

  1. Normalization by variance stabilizing transformation
  2. PCA dimensionality reduction
  3. UMAP dimensionality reduction
  4. Clustering (FindNeighbors + FindClusters)

```{r normalization}

se <- se %>% SCTransform() %>%
  RunPCA() %>%
  RunUMAP(reduction = "pca", dims = 1:30) %>%
  FindNeighbors(reduction = "pca", dims = 1:30) %>%
  FindClusters()

```

UMAP without integration

```{r dimplot, fig.width = 8, fig.height = 4}

p1 <- DimPlot(se, group.by = "sample_id")
p2 <- DimPlot(se)
p1 - p2

```


Align images 

```{r}
se <- LoadImages(se, time.resolve = FALSE)

# Apply simple rigid transformations to images
se <- WarpImages(se, transforms = list("1" = list(angle = 60), 
                                       "2" = list(shift.x = -50, shift.y = 50), 
                                       "5" = list(angle = -10), 
                                       "6" = list(angle = 20), 
                                       "7" = list(angle = 25)))
```

```{r, fig.width=8, fig.height=6}

ImagePlot(se, method="raster")

```


## Harmony
***

Integration with harmony

```{r harmony, fig.width=10, fig.height=10}

library(harmony)
se <- RunHarmony(se, group.by.vars = "date", dims.use = 1:50, assay.use = "SCT")
se <- RunUMAP(se, reduction = "harmony", dims = 1:50, reduction.name = "umap.harmony")
se <- FindNeighbors(se, dims = 1:50, reduction = "harmony")
se <- se  %>% FindClusters(resolution = 0.8)
se <- RunUMAP(se, dims = 1:50, reduction = "harmony", n.components = 3, reduction.name = "umap.3d")

```

Visualize RGB encoded spots

```{r rgb, fig.width=12, fig.height=3}

p <- ST.DimPlot(se, dims = -c(1:3), reduction = "umap.3d", blend = T, pt.size = 1.2, show.sb = F, ncol = 7, custom.theme = theme(plot.margin = margin(0,0,0,0, "cm"), strip.text = element_blank()))
p
  
```

```{r umap, fig.width=14, fig.height=5}

p <- DimPlot(se, reduction = "umap.harmony", split.by = "date")
p

```

Relative composition of clusters by sample

```{r umap, fig.width=14, fig.height=5}

gg <- data.frame(cluster = se$seurat_clusters, experiment = se$date)
gg <- gg %>%
  group_by(experiment, cluster) %>%
  summarize(count = n()) %>%
  group_by(experiment) %>%
  mutate(Freq = count/sum(count))

p <- ggplot(gg) +
  geom_bar(aes(factor(experiment, levels = c("201016", "200925", "200513")), Freq, fill = cluster), stat = "identity", color = "black") +
  coord_flip() +
  scale_y_reverse() +
  theme_minimal()
p

```

```{r qc, fig.width=14, fig.height=10}

gg <- se@meta.data
p1 <- ggplot() +
  geom_violin(data = gg, aes(sample_id, nFeature_RNA, color = sample_id, fill = sample_id), alpha = 0.6) +
  geom_boxplot(data = gg, aes(sample_id, nFeature_RNA, color = sample_id, fill = sample_id), alpha = 1, color = "black", width = 0.3) +
  facet_grid(~date, scales = "free_x", space='free') +
  theme_bw() +
  labs(y = "unique genes", x = "") +
  guides(fill = F, color = F)
p2 <- ggplot() +
  geom_violin(data = gg, aes(sample_id, nCount_RNA, color = sample_id, fill = sample_id), alpha = 0.6) +
  geom_boxplot(data = gg, aes(sample_id, nCount_RNA, color = sample_id, fill = sample_id), alpha = 1, color = "black", width = 0.3) +
  scale_y_continuous(breaks = c(0, 5e3, 10e3, 15e3, 20e3, 25e3)) +
  facet_grid(~date, scales = "free_x", space='free') +
  theme_bw() +
  labs(y = "UMIs", x = "") +
  guides(fill = F, color = F)
p3 <- p1 / p2
p3

```

## Date
***

```{r date}

date()

```

## Session
***

```{r}

sessionInfo()

```



